---
title: "03-nobones-plotting"
author: "Kathleen Durkin"
date: "2025-12-14"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

# Sequencing runs plots


# Setup and Data Loading


```{r load-sequencing-summary, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(patchwork)  # for combining plots
library(viridis)    # for color palettes

# Define color scheme
age_colors <- c(
  "Modern (2019)" = "#27ae60",
  "Mid-age (1960)" = "#3498db",
  "Old (1880s)" = "#e74c3c"
)

# Load sequencing summary data
seq_summary <- read_csv("../../data/E_tourneforti_sequencing_summaries.csv")

# Clean up and add derived columns
seq_summary <- seq_summary %>%
  mutate(
    # Create cleaner run labels
    Run_Label = paste0("G", Group, "L", Library),
    
    # Combine pre/post runs for the same library
    Library_ID = paste0("Group", Group, "_Library", Library),
    
    # Calculate pass rate
    Pass_Rate = Reads.Called.Pass.M / Reads.Generated.M * 100,
    
    # Calculate bases pass rate
    Bases_Pass_Rate = Bases.Called.Pass.Gb / Estimated.Bases.Gb * 100,
    
    # Throughput rate (Gb per hour)
    Throughput_Rate = Estimated.Bases.Gb / Run.time.hr,
    
    # Standardize sequencer names
    Sequencer = case_when(
      Sequencer %in% c("MK1D", "Mk1D") ~ "MinION Mk1D",
      Sequencer %in% c("Mk1B", "MK1B") ~ "MinION Mk1B",
      TRUE ~ Sequencer
    ),
    
    # Add age group based on Group number
    Age_Group = case_when(
      Group == 1 ~ "Modern (2019)",
      Group == 2 ~ "Mid-age (1960)",
      Group %in% c(3, 4) ~ "Old (1880s)",
      TRUE ~ "Unknown"
    ),
    Age_Group = factor(
      Age_Group,
      levels = c("Old (1880s)", "Mid-age (1960)", "Modern (2019)")
    )
  )

# ---- Aggregate by Group + Library ----
seq_summary_agg <- seq_summary_clean %>%
  select(-Estimated.N50.b) %>%
  group_by(Group, Library) %>%
  summarise(
    # IDs
    Library_ID = paste0("Group", Group, "_Lib", Library),
    Age_Group = first(Age_Group),
    Sequencer = first(Sequencer),
    
    # Accumulating quantities → SUM
    Estimated.Bases.Gb = sum(Estimated.Bases.Gb, na.rm = TRUE),
    Reads.Generated.M  = sum(Reads.Generated.M, na.rm = TRUE),
    Reads.Called.Pass.M = sum(Reads.Called.Pass.M, na.rm = TRUE),
    Bases.Called.Pass.Gb = sum(Bases.Called.Pass.Gb, na.rm = TRUE),
    Run.time.hr = sum(Run.time.hr, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  mutate(
    # Recalculate rates AFTER aggregation
    Pass_Rate = Reads.Called.Pass.M / Reads.Generated.M * 100,
    Bases_Pass_Rate = Bases.Called.Pass.Gb / Estimated.Bases.Gb * 100,
    Throughput_Rate = Estimated.Bases.Gb / Run.time.hr,
    Age_Group = factor(
      Age_Group,
      levels = c("Old (1880s)", "Mid-age (1960)", "Modern (2019)"))
  ) %>%
  distinct()

# Display the cleaned data
seq_summary %>%
  select(Sequencing.Run, Group, Library, Library_ID, Estimated.Bases.Gb, Reads.Generated.M, 
         Estimated.N50.b, Pass_Rate, Sequencer) %>%
  knitr::kable(caption = "Sequencing Run Summary", digits = 2)
```


# Figure 1: Total Output by Sequencing Run


```{r fig-total-output, eval=TRUE, fig.width=10, fig.height=6}
# Bar plot of total bases by run
p1 <- ggplot(seq_summary_agg, aes(x = reorder(Library_ID, Estimated.Bases.Gb), 
                               y = Estimated.Bases.Gb,
                               fill = Age_Group)) +
geom_col() +
  coord_flip() +
  labs(
    title = "Total Sequencing Output by Run",
    subtitle = "E. tourneforti Nanopore sequencing",
    x = "Sequencing Run",
    y = "Estimated Bases (Gb)",
    fill = "Age Group"
  ) +
  scale_fill_manual(values = c(
    "Modern (2019)" = "#27ae60",
    "Mid-age (1960)" = "#3498db",
    "Old (1880s)" = "#e74c3c"
  )) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = "bottom"
  )

print(p1)

ggsave("../output/03-nobones-plotting/fig_total_output_by_run.png", 
       p1, width = 10, height = 6, dpi = 150)
```


# Figure 2: Reads Generated vs Bases (Quality Assessment)


```{r fig-reads-vs-bases, eval=TRUE, fig.width=8, fig.height=6}
p2 <- ggplot(seq_summary, aes(x = Reads.Generated.M, y = Estimated.Bases.Gb)) +
  geom_point(aes(color = Age_Group, size = Estimated.N50.b), alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "gray40", linetype = "dashed") +
  geom_text(aes(label = Run_Label), vjust = -1, hjust = 0.5, size = 3) +
  labs(
    title = "Reads vs Total Bases Output",
    subtitle = "Point size indicates N50 read length",
    x = "Reads Generated (millions)",
    y = "Estimated Bases (Gb)",
    color = "Age Group",
    size = "N50 (bp)"
  ) +
  scale_color_manual(values = c(
    "Modern (2019)" = "#27ae60",
    "Mid-age (1960)" = "#3498db",
    "Old (1880s)" = "#e74c3c"
  )) +
  theme_minimal() +
  theme(legend.position = "right")

print(p2)

ggsave("../output/03-nobones-plotting/fig_reads_vs_bases.png", 
       p2, width = 8, height = 6, dpi = 150)
```


# Figure 3: N50 Read Length by Run


```{r fig-n50, eval=TRUE, fig.width=10, fig.height=5}
p3 <- ggplot(seq_summary, aes(x = reorder(Sequencing.Run, Estimated.N50.b), 
                               y = Estimated.N50.b,
                               fill = Age_Group)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 500, linetype = "dashed", color = "gray50") +
  annotate("text", x = 0.5, y = 520, label = "500 bp", size = 3, hjust = 0) +
  labs(
    title = "N50 Read Length by Sequencing Run",
    subtitle = "Higher N50 indicates longer reads",
    x = "Sequencing Run",
    y = "Estimated N50 (bp)",
    fill = "Age Group"
  ) +
  scale_fill_manual(values = c(
    "Modern (2019)" = "#27ae60",
    "Mid-age (1960)" = "#3498db",
    "Old (1880s)" = "#e74c3c"
  )) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p3)

ggsave("../output/03-nobones-plotting/fig_n50_by_run.png", 
       p3, width = 10, height = 5, dpi = 150)
```


# Figure 4: Pass vs Fail Reads


```{r fig-pass-fail, eval=TRUE, fig.width=10, fig.height=6}
# Reshape data for stacked bar
pass_fail_data <- seq_summary %>%
  select(Sequencing.Run, Age_Group, Reads.Called.Pass.M, Reads.Called.Fail.M) %>%
  pivot_longer(
    cols = c(Reads.Called.Pass.M, Reads.Called.Fail.M),
    names_to = "Read_Type",
    values_to = "Reads_M"
  ) %>%
  mutate(
    Read_Type = ifelse(Read_Type == "Reads.Called.Pass.M", "Pass", "Fail")
  )

p4 <- ggplot(pass_fail_data, aes(x = reorder(Sequencing.Run, Reads_M), 
                                  y = Reads_M,
                                  fill = Read_Type)) +
  geom_col(position = "stack") +
  coord_flip() +
  labs(
    title = "Pass vs Fail Reads by Sequencing Run",
    x = "Sequencing Run",
    y = "Reads (millions)",
    fill = "Read Quality"
  ) +
  scale_fill_manual(values = c("Pass" = "#2ecc71", "Fail" = "#e74c3c")) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p4)

ggsave("../output/03-nobones-plotting/fig_pass_fail_reads.png", 
       p4, width = 10, height = 6, dpi = 150)
```


# Figure 5: Pass Rate Comparison


```{r fig-pass-rate, eval=TRUE, fig.width=10, fig.height=5}
p5 <- ggplot(seq_summary, aes(x = reorder(Sequencing.Run, Pass_Rate), 
                               y = Pass_Rate,
                               fill = Age_Group)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 90, linetype = "dashed", color = "gray50") +
  labs(
    title = "Read Pass Rate by Sequencing Run",
    subtitle = "Percentage of reads passing quality filters",
    x = "Sequencing Run",
    y = "Pass Rate (%)",
    fill = "Age Group"
  ) +
  scale_fill_manual(values = c(
    "Modern (2019)" = "#27ae60",
    "Mid-age (1960)" = "#3498db",
    "Old (1880s)" = "#e74c3c"
  )) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p5)

ggsave("../output/03-nobones-plotting/fig_pass_rate.png", 
       p5, width = 10, height = 5, dpi = 150)
```


# Figure 6: Throughput Rate (Gb/hr)


```{r fig-throughput, eval=TRUE, fig.width=10, fig.height=5}
p6 <- ggplot(seq_summary, aes(x = reorder(Sequencing.Run, Throughput_Rate), 
                               y = Throughput_Rate,
                               fill = Sequencer)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Sequencing Throughput Rate",
    subtitle = "Output normalized by run time",
    x = "Sequencing Run",
    y = "Throughput (Gb/hr)",
    fill = "Sequencer"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p6)

ggsave("../output/03-nobones-plotting/fig_throughput_rate.png", 
       p6, width = 10, height = 5, dpi = 150)
```


# Figure 7: Summary by Age Group (Aggregated)


```{r fig-age-group-summary, eval=TRUE, fig.width=12, fig.height=8}
# Calculate totals for age group labels
age_totals <- seq_summary %>%
  group_by(Age_Group) %>%
  summarise(
    Total_Reads_M = sum(Reads.Generated.M),
    Total_Bases_Gb = sum(Estimated.Bases.Gb),
    .groups = "drop"
  )

# Panel A: Total reads by age group (STACKED by Library_ID)
p7a <- ggplot(seq_summary, aes(x = Age_Group, y = Reads.Generated.M, fill = Age_Group)) +
  geom_col(aes(alpha = Library_ID), color = "white", linewidth = 0.3) +
  geom_text(data = age_totals, 
            aes(x = Age_Group, y = Total_Reads_M, label = sprintf("%.1f M", Total_Reads_M)),
            vjust = -0.5, inherit.aes = FALSE, size = 3.5) +
  # Add library labels inside bars
  geom_text(aes(label = Library_ID), 
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white", fontface = "bold") +
  labs(title = "A) Total Reads", x = "", y = "Total Reads (millions)") +
  scale_fill_manual(values = age_colors) +
  scale_alpha_manual(values = seq(0.6, 1, length.out = 5), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))

# Panel B: Total bases by age group (STACKED by Library_ID)
p7b <- ggplot(seq_summary, aes(x = Age_Group, y = Estimated.Bases.Gb, fill = Age_Group)) +
  geom_col(aes(alpha = Library_ID), color = "white", linewidth = 0.3) +
  geom_text(data = age_totals, 
            aes(x = Age_Group, y = Total_Bases_Gb, label = sprintf("%.2f Gb", Total_Bases_Gb)),
            vjust = -0.5, inherit.aes = FALSE, size = 3.5) +
  # Add library labels inside bars
  geom_text(aes(label = Library_ID), 
            position = position_stack(vjust = 0.5), 
            size = 3, color = "white", fontface = "bold") +
  labs(title = "B) Total Bases", x = "", y = "Total Bases (Gb)") +
  scale_fill_manual(values = age_colors) +
  scale_alpha_manual(values = seq(0.6, 1, length.out = 5), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))

# Panel C: N50 by Library (individual bars, colored by Age Group)
p7c <- ggplot(seq_summary, aes(x = Library_ID, y = Estimated.N50.b, fill = Age_Group)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f", Estimated.N50.b)), vjust = -0.5, size = 3) +
  labs(title = "C) N50 Read Length", x = "", y = "N50 (bp)") +
  scale_fill_manual(values = age_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Panel D: Pass rate by Library (individual bars, colored by Age Group)
p7d <- ggplot(seq_summary, aes(x = Library_ID, y = Pass_Rate, fill = Age_Group)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", Pass_Rate)), vjust = -0.5, size = 3) +
  geom_hline(yintercept = 90, linetype = "dashed", color = "gray50", alpha = 0.7) +
  labs(title = "D) Read Pass Rate", x = "", y = "Pass Rate (%)") +
  scale_fill_manual(values = age_colors) +
  scale_y_continuous(limits = c(0, 105), expand = expansion(mult = c(0, 0))) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Create a shared legend
legend_plot <- ggplot(seq_summary, aes(x = Library_ID, y = Pass_Rate, fill = Age_Group)) +
  geom_col() +
  scale_fill_manual(values = age_colors) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_blank())

# Extract legend
library(cowplot)
shared_legend <- get_legend(legend_plot)

# Combine panels
p7_combined <- (p7a | p7b) / (p7c | p7d) +
  plot_annotation(
    title = "Sequencing Summary by Specimen Age Group",
    subtitle = "E. tourneforti archival coral specimens"
  )

print(p7_combined)

ggsave("../output/05-specimen-coverage/fig_age_group_summary.png", 
       p7_combined, width = 12, height = 8, dpi = 150)
```


# Figure 8: Run Time vs Output


```{r fig-runtime-output, eval=TRUE, fig.width=8, fig.height=6}
p8 <- ggplot(seq_summary, aes(x = Run.time.hr, y = Estimated.Bases.Gb)) +
  geom_point(aes(color = Age_Group, shape = Sequencer), size = 4, alpha = 0.8) +
  geom_text(aes(label = Run_Label), vjust = -1, size = 3) +
  labs(
    title = "Run Time vs Sequencing Output",
    x = "Run Time (hours)",
    y = "Estimated Bases (Gb)",
    color = "Age Group",
    shape = "Sequencer"
  ) +
  scale_color_manual(values = c(
    "Modern (2019)" = "#27ae60",
    "Mid-age (1960)" = "#3498db",
    "Old (1880s)" = "#e74c3c"
  )) +
  theme_minimal() +
  theme(legend.position = "right")

print(p8)

ggsave("../output/03-nobones-plotting/fig_runtime_vs_output.png", 
       p8, width = 8, height = 6, dpi = 150)
```


# Figure 9: Bases Pass vs Fail (Stacked)


```{r fig-bases-pass-fail, eval=TRUE, fig.width=10, fig.height=6}
# Reshape for bases
bases_data <- seq_summary %>%
  select(Sequencing.Run, Age_Group, Bases.Called.Pass.Gb, Bases.Called.Fail.Gb) %>%
  pivot_longer(
    cols = c(Bases.Called.Pass.Gb, Bases.Called.Fail.Gb),
    names_to = "Base_Type",
    values_to = "Bases_Gb"
  ) %>%
  mutate(
    Base_Type = ifelse(Base_Type == "Bases.Called.Pass.Gb", "Pass", "Fail")
  )

p9 <- ggplot(bases_data, aes(x = reorder(Sequencing.Run, Bases_Gb), 
                              y = Bases_Gb,
                              fill = Base_Type)) +
  geom_col(position = "stack") +
  coord_flip() +
  labs(
    title = "Pass vs Fail Bases by Sequencing Run",
    x = "Sequencing Run",
    y = "Bases (Gb)",
    fill = "Base Quality"
  ) +
  scale_fill_manual(values = c("Pass" = "#2ecc71", "Fail" = "#e74c3c")) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p9)

ggsave("../output/03-nobones-plotting/fig_bases_pass_fail.png", 
       p9, width = 10, height = 6, dpi = 150)
```


# Summary Table


```{r summary-table, eval=TRUE}
# Create a nice summary table
summary_table <- seq_summary %>%
  select(
    Run = Sequencing.Run,
    Group = Age_Group,
    `Bases (Gb)` = Estimated.Bases.Gb,
    `Reads (M)` = Reads.Generated.M,
    `N50 (bp)` = Estimated.N50.b,
    `Pass Rate (%)` = Pass_Rate,
    `Run Time (hr)` = Run.time.hr,
    Sequencer
  ) %>%
  arrange(desc(`Bases (Gb)`))

knitr::kable(summary_table, 
             caption = "Sequencing Run Summary Statistics",
             digits = 2)

# Age group summary table
age_summary %>%
  mutate(
    `Bases (Gb)` = round(Total_Bases_Gb, 2),
    `Reads (M)` = round(Total_Reads_M, 2),
    `Mean N50 (bp)` = round(Mean_N50, 0),
    `Mean Pass Rate (%)` = round(Mean_Pass_Rate, 1),
    `Total Run Time (hr)` = round(Total_Run_Time_hr, 1)
  ) %>%
  select(Age_Group, n_runs, `Bases (Gb)`, `Reads (M)`, 
         `Mean N50 (bp)`, `Mean Pass Rate (%)`, `Total Run Time (hr)`) %>%
  knitr::kable(caption = "Summary Statistics by Age Group")
```


# Statistical Comparisons (Optional)


```{r stats-comparison, eval=TRUE}
# Compare N50 between age groups
cat("=== N50 Comparison by Age Group ===\n")
seq_summary %>%
  group_by(Age_Group) %>%
  summarise(
    mean_N50 = mean(Estimated.N50.b),
    sd_N50 = sd(Estimated.N50.b),
    n = n(),
    .groups = "drop"
  ) %>%
  print()

# Simple correlation between bases and reads
cat("\n=== Correlation: Bases vs Reads ===\n")
cor_test <- cor.test(seq_summary$Estimated.Bases.Gb, seq_summary$Reads.Generated.M)
cat("Pearson r =", round(cor_test$estimate, 3), "\n")
cat("p-value =", format(cor_test$p.value, scientific = TRUE), "\n")
```









# Plot modifications by specimen age

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)  # for combining plots
```

```{r}

# ===== STEP 1: Find all modkit summary files =====
modkit_files <- list.files(
  path = "../../", 
  pattern = "modkit_summary_.*\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)

cat("Found", length(modkit_files), "modkit summary files\n")

# ===== STEP 2: Read and parse all files =====
# Read each file and extract modification data
modkit_data <- map_dfr(modkit_files, function(filepath) {
  # Read as tab-delimited key-value pairs
  data <- read_tsv(filepath, col_names = c("key", "value"), show_col_types = FALSE)
  
  # Convert to wide format
  data_wide <- data %>%
    pivot_wider(names_from = key, values_from = value)
  
  # Extract barcode from filename
  barcode <- str_extract(basename(filepath), "\\d+(?=\\.tsv)")
  
  # Extract modification percentages
  data_wide %>%
    mutate(
      barcode = as.numeric(barcode),
      # Cytosine modifications
      C_5mC_pct = as.numeric(C_pass_frac_modified_m) * 100,
      C_5hmC_pct = as.numeric(C_pass_frac_modified_h) * 100,
      C_unmod_pct = as.numeric(C_pass_frac_unmodified) * 100,
      # Adenine modifications
      A_6mA_pct = as.numeric(A_pass_frac_modified_a) * 100,
      A_unmod_pct = as.numeric(A_pass_frac_unmodified) * 100
    ) %>%
    select(barcode, ends_with("_pct"))
})


```


## Load and Merge All Metadata

```{r merge-all-metadata}
# 1. Load sequencing composition (barcode -> catalog number, year, group/library)
composition <- read_csv("../../data/E_tourneforti_sequencing_composition.csv", 
                       show_col_types = FALSE)

# 2. Load sample metadata (catalog number -> DIN, other DNA quality metrics)
samples <- read_csv("../../data/E_tourneforti_E_tourneforti_samples.csv",
                   show_col_types = FALSE) %>%
  select(Catalog.Number, Year.Collected, Tapestation.DIN, 
         Tapestation.Conc.ngul, Tapestation.Perc.300up)
samples$Catalog.Number <- as.double(gsub("USNM ", "", samples$Catalog.Number))

# 3. Load sequencing summaries (group/library -> N50)
seq_summaries <- read_csv("../../data/E_tourneforti_sequencing_summaries.csv",
                         show_col_types = FALSE) %>%
  select(Group, Library, Estimated.N50.b, Estimated.Bases.Gb, 
         Reads.Generated.M, Sequencer)
seq_summaries$Group <- paste0("Group_", seq_summaries$Group)
seq_summaries$Library <- paste0("Library_", seq_summaries$Library)

# 4. Join all metadata together
metadata <- composition %>%
  # First join with samples to get DIN
  left_join(samples, by = "Catalog.Number") %>%
  # Then join with sequencing summaries to get N50
  left_join(seq_summaries, by = c("Group", "Library"))

# 5. Join with modkit data
plot_data <- modkit_data %>%
  left_join(metadata, by = c("barcode" = "Barcode")) %>%
  arrange(Year.Collected.x) %>%
  # Use Year.Collected from composition file (Year.Collected.x)
  mutate(Year.Collected = Year.Collected.x) %>%
  select(-Year.Collected.x, -Year.Collected.y)

# Display merged data table
plot_data %>%
  select(barcode, Year.Collected, Catalog.Number, 
         C_5mC_pct, C_5hmC_pct, A_6mA_pct, 
         Tapestation.DIN, Estimated.N50.b) %>%
  kable(digits = 2, caption = "Modification percentages with DIN and N50") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 11) %>%
  scroll_box(width = "100%", height = "400px")
```

## Reshape Data for Plotting

```{r reshape-data}
plot_data_long <- plot_data %>%
  pivot_longer(
    cols = ends_with("_pct"),
    names_to = "modification",
    values_to = "percentage"
  ) %>%
  filter(!is.na(percentage)) %>%
  separate(modification, into = c("base", "mod_type", "pct"), sep = "_") %>%
  select(-pct) %>%
  mutate(
    mod_label = case_when(
      mod_type == "5mC" ~ "5mC (methylcytosine)",
      mod_type == "5hmC" ~ "5hmC (hydroxymethylcytosine)", 
      mod_type == "6mA" ~ "6mA (methyladenine)",
      mod_type == "unmod" ~ "Unmodified",
      TRUE ~ mod_type
    ),
    full_label = paste0(base, " - ", mod_label),
    # Create age categories
    age_category = case_when(
      Year.Collected >= 2000 ~ "Modern (<25 years)",
      Year.Collected >= 1960 ~ "Mid-age (1960s)",
      TRUE ~ "Historic (>100 years)"
    )
  )
```

# Modifications vs. Sample Age

```{r}
plot_mod_data <- plot_data %>%
  filter(mod_type %in% c("5mC", "5hmC", "6mA")) %>%
  group_by(Catalog.Number, mod_type) %>%
  summarise(
    percentage = mean(percentage, na.rm = TRUE),
    Year.Collected = first(Year.Collected),
    age_category = first(age_category),
    .groups = "drop"
  ) %>%
  mutate(
    # order specimens by collection year
    Catalog.Number = fct_reorder(
      as.factor(Catalog.Number),
      Year.Collected
    )
  )

age_colors <- c(
  "Historic (>100 years)" = "#e74c3c",
  "Mid-age (1960s)" = "#3498db",
  "Modern (<25 years)" = "#27ae60"
)

plot_mod_bar <- function(mod) {
  
  ggplot(
    plot_mod_data %>% filter(mod_type == mod),
    aes(
      x = Catalog.Number,
      y = percentage,
      fill = age_category
    )
  ) +
    geom_col(width = 0.8) +
    geom_text(
      aes(label = Year.Collected),
      vjust = -0.4,
      size = 3
    ) +
    scale_fill_manual(values = age_colors) +
    labs(
      title = paste0(mod, " Modification per Specimen"),
      x = "Specimen (Catalog Number)",
      y = "% Modified"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.title = element_blank()
    )
}


p_5mC  <- plot_mod_bar("5mC")
p_5hmC <- plot_mod_bar("5hmC")
p_6mA  <- plot_mod_bar("6mA")

p_5mC
p_5hmC
p_6mA

p_combined <- (p_5mC | p_5hmC | p_6mA) +
  plot_annotation(
    title = "DNA Modification Levels Across Specimens",
    subtitle = "Bars colored by specimen age category; labels indicate year collected"
  ) &
  theme(legend.position = "bottom")

ggsave(
  filename = "../output/03-nobones-plotting/DNA_modification_levels_by_specimen.png",
  plot = p_combined,
  width = 14,
  height = 5,
  dpi = 300
)

```


```{r plot-by-age, fig.height=9}
p_age <- plot_data_long %>%
  filter(mod_type != "unmod") %>%
  ggplot(aes(x = Year.Collected, y = percentage, color = base)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
  facet_wrap(~mod_label, scales = "free_y", ncol = 1) +
  scale_color_manual(
    values = c("C" = "#E41A1C", "A" = "#377EB8"),
    name = "Base"
  ) +
  labs(
    title = "DNA Modifications by Sample Age",
    subtitle = "Linear regression with 95% confidence interval",
    x = "Year Collected",
    y = "Percentage (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )

print(p_age)
```

# Modifications vs. DNA Integrity (DIN)

The DNA Integrity Number (DIN) is a measure of DNA quality from TapeStation analysis. Higher DIN values indicate better DNA integrity.

```{r plot-by-din, fig.height=9}
p_din <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage, color = age_category)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
  facet_wrap(~mod_label, scales = "free_y", ncol = 1) +
  scale_color_brewer(palette = "Set1", name = "Age Category") +
  labs(
    title = "DNA Modifications vs. DNA Integrity (DIN)",
    subtitle = "Linear regression with 95% confidence interval",
    x = "DNA Integrity Number (DIN)",
    y = "Percentage (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )

print(p_din)
```

# Modifications vs. Sequencing N50

N50 is a measure of read length quality - the length at which 50% of bases are in reads of that length or longer.

```{r plot-by-n50, fig.height=9}
p_n50 <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage, color = age_category)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
  facet_wrap(~mod_label, scales = "free_y", ncol = 1) +
  scale_color_brewer(palette = "Set1", name = "Age Category") +
  labs(
    title = "DNA Modifications vs. Sequencing N50",
    subtitle = "Linear regression with 95% confidence interval",
    x = "N50 (bases)",
    y = "Percentage (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )

print(p_n50)
```

# Combined Comparison Plots

## 5mC across all three variables

```{r combined-5mc, fig.height=4, fig.width=12}
p_5mc_age <- plot_data_long %>%
  filter(mod_type == "5mC") %>%
  ggplot(aes(x = Year.Collected, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. Age", x = "Year Collected", y = "5mC (%)") +
  theme_bw()

p_5mc_din <- plot_data_long %>%
  filter(mod_type == "5mC", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. DIN", x = "DIN", y = "5mC (%)") +
  theme_bw()

p_5mc_n50 <- plot_data_long %>%
  filter(mod_type == "5mC", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. N50", x = "N50 (bases)", y = "5mC (%)") +
  theme_bw()

p_5mc_age + p_5mc_din + p_5mc_n50
```

## 5hmC across all three variables

```{r combined-5hmc, fig.height=4, fig.width=12}

p_5hmc_age <- plot_data_long %>%
  filter(mod_type == "5hmC") %>%
  ggplot(aes(x = Year.Collected, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. Age", x = "Year Collected", y = "5hmC (%)") +
  theme_bw()

p_5hmc_din <- plot_data_long %>%
  filter(mod_type == "5hmC", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5hmC vs. DIN", x = "DIN", y = "5hmC (%)") +
  theme_bw()

p_5hmc_n50 <- plot_data_long %>%
  filter(mod_type == "5hmC", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5hmC vs. N50", x = "N50 (bases)", y = "5hmC (%)") +
  theme_bw()

p_5hmc_age + p_5hmc_din + p_5hmc_n50
```



## 6mA across all three variables

```{r combined-6ma, fig.height=4, fig.width=12}
p_6ma_age <- plot_data_long %>%
  filter(mod_type == "6mA") %>%
  ggplot(aes(x = Year.Collected, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#377EB8") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8") +
  labs(title = "6mA vs. Age", x = "Year Collected", y = "6mA (%)") +
  theme_bw()

p_6ma_din <- plot_data_long %>%
  filter(mod_type == "6mA", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#377EB8") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8") +
  labs(title = "6mA vs. DIN", x = "DIN", y = "6mA (%)") +
  theme_bw()

p_6ma_n50 <- plot_data_long %>%
  filter(mod_type == "6mA", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#377EB8") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8") +
  labs(title = "6mA vs. N50", x = "N50 (bases)", y = "6mA (%)") +
  theme_bw()

p_6ma_age + p_6ma_din + p_6ma_n50
```

# Correlation Analysis

## Correlations with Year Collected

```{r cor-age}
cor_age <- plot_data_long %>%
  filter(mod_type != "unmod") %>%
  group_by(mod_label) %>%
  summarise(
    r = cor(Year.Collected, percentage, use = "complete.obs"),
    p_value = cor.test(Year.Collected, percentage)$p.value,
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.0001 ~ "****",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

cor_age %>%
  kable(digits = 4, caption = "Correlation with Year Collected",
        col.names = c("Modification", "r", "P-value", "N", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Correlations with DIN

```{r cor-din}
cor_din <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Tapestation.DIN)) %>%
  group_by(mod_label) %>%
  summarise(
    r = cor(Tapestation.DIN, percentage, use = "complete.obs"),
    p_value = cor.test(Tapestation.DIN, percentage)$p.value,
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.0001 ~ "****",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

cor_din %>%
  kable(digits = 4, caption = "Correlation with DIN",
        col.names = c("Modification", "r", "P-value", "N", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Correlations with N50

```{r cor-n50}
cor_n50 <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Estimated.N50.b)) %>%
  group_by(mod_label) %>%
  summarise(
    r = cor(Estimated.N50.b, percentage, use = "complete.obs"),
    p_value = cor.test(Estimated.N50.b, percentage)$p.value,
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.0001 ~ "****",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

cor_n50 %>%
  kable(digits = 4, caption = "Correlation with N50",
        col.names = c("Modification", "r", "P-value", "N", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Summary Statistics by Age Category

```{r summary-by-age-cat}
summary_by_age <- plot_data_long %>%
  filter(mod_type != "unmod") %>%
  group_by(age_category, mod_label) %>%
  summarise(
    n = n(),
    mean_pct = mean(percentage, na.rm = TRUE),
    sd_pct = sd(percentage, na.rm = TRUE),
    mean_din = mean(Tapestation.DIN, na.rm = TRUE),
    mean_n50 = mean(Estimated.N50.b, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(age_category, desc(mean_pct))

summary_by_age %>%
  kable(digits = 2, 
        caption = "Summary statistics by age category",
        col.names = c("Age Category", "Modification", "N", 
                      "Mean %", "SD %", "Mean DIN", "Mean N50")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  collapse_rows(columns = 1, valign = "top")
```

# Interpretation Notes

## Key Findings

- **5mC (methylcytosine)**: Standard DNA methylation mark in eukaryotes
- **5hmC (hydroxymethylcytosine)**: Oxidation product of 5mC, potentially unstable in old samples
- **6mA (N6-methyladenine)**: Uncommon in eukaryotes, may indicate:
  - Contamination (bacterial/fungal DNA)
  - Age-related degradation artifacts
  - Sequencing/basecalling artifacts in degraded DNA

## Relationships to consider

- **Age effects**: Are modifications changing with sample age?
- **DIN effects**: Does DNA integrity affect apparent modification levels?
- **N50 effects**: Does sequencing quality correlate with modification detection?

High 6mA levels in older, lower-quality samples warrant careful interpretation.

