---
title: "03-nobones-plotting"
author: "Kathleen Durkin"
date: "2025-12-14"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---


Plot modifications by specimen age

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)  # for combining plots
```

```{r}

# ===== STEP 1: Find all modkit summary files =====
modkit_files <- list.files(
  path = "../../", 
  pattern = "modkit_summary_.*\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)

cat("Found", length(modkit_files), "modkit summary files\n")

# ===== STEP 2: Read and parse all files =====
# Read each file and extract modification data
modkit_data <- map_dfr(modkit_files, function(filepath) {
  # Read as tab-delimited key-value pairs
  data <- read_tsv(filepath, col_names = c("key", "value"), show_col_types = FALSE)
  
  # Convert to wide format
  data_wide <- data %>%
    pivot_wider(names_from = key, values_from = value)
  
  # Extract barcode from filename
  barcode <- str_extract(basename(filepath), "\\d+(?=\\.tsv)")
  
  # Extract modification percentages
  data_wide %>%
    mutate(
      barcode = as.numeric(barcode),
      # Cytosine modifications
      C_5mC_pct = as.numeric(C_pass_frac_modified_m) * 100,
      C_5hmC_pct = as.numeric(C_pass_frac_modified_h) * 100,
      C_unmod_pct = as.numeric(C_pass_frac_unmodified) * 100,
      # Adenine modifications
      A_6mA_pct = as.numeric(A_pass_frac_modified_a) * 100,
      A_unmod_pct = as.numeric(A_pass_frac_unmodified) * 100
    ) %>%
    select(barcode, ends_with("_pct"))
})


```


## Load and Merge All Metadata

```{r merge-all-metadata}
# 1. Load sequencing composition (barcode -> catalog number, year, group/library)
composition <- read_csv("../../data/E_tourneforti_sequencing_composition.csv", 
                       show_col_types = FALSE)

# 2. Load sample metadata (catalog number -> DIN, other DNA quality metrics)
samples <- read_csv("../../data/E_tourneforti_E_tourneforti_samples.csv",
                   show_col_types = FALSE) %>%
  select(Catalog.Number, Year.Collected, Tapestation.DIN, 
         Tapestation.Conc.ngul, Tapestation.Perc.300up)
samples$Catalog.Number <- as.double(gsub("USNM ", "", samples$Catalog.Number))

# 3. Load sequencing summaries (group/library -> N50)
seq_summaries <- read_csv("../../data/E_tourneforti_sequencing_summaries.csv",
                         show_col_types = FALSE) %>%
  select(Group, Library, Estimated.N50.b, Estimated.Bases.Gb, 
         Reads.Generated.M, Sequencer)
seq_summaries$Group <- paste0("Group_", seq_summaries$Group)
seq_summaries$Library <- paste0("Library_", seq_summaries$Library)

# 4. Join all metadata together
metadata <- composition %>%
  # First join with samples to get DIN
  left_join(samples, by = "Catalog.Number") %>%
  # Then join with sequencing summaries to get N50
  left_join(seq_summaries, by = c("Group", "Library"))

# 5. Join with modkit data
plot_data <- modkit_data %>%
  left_join(metadata, by = c("barcode" = "Barcode")) %>%
  arrange(Year.Collected.x) %>%
  # Use Year.Collected from composition file (Year.Collected.x)
  mutate(Year.Collected = Year.Collected.x) %>%
  select(-Year.Collected.x, -Year.Collected.y)

# Display merged data table
plot_data %>%
  select(barcode, Year.Collected, Catalog.Number, 
         C_5mC_pct, C_5hmC_pct, A_6mA_pct, 
         Tapestation.DIN, Estimated.N50.b) %>%
  kable(digits = 2, caption = "Modification percentages with DIN and N50") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 11) %>%
  scroll_box(width = "100%", height = "400px")
```

## Reshape Data for Plotting

```{r reshape-data}
plot_data_long <- plot_data %>%
  pivot_longer(
    cols = ends_with("_pct"),
    names_to = "modification",
    values_to = "percentage"
  ) %>%
  filter(!is.na(percentage)) %>%
  separate(modification, into = c("base", "mod_type", "pct"), sep = "_") %>%
  select(-pct) %>%
  mutate(
    mod_label = case_when(
      mod_type == "5mC" ~ "5mC (methylcytosine)",
      mod_type == "5hmC" ~ "5hmC (hydroxymethylcytosine)", 
      mod_type == "6mA" ~ "6mA (methyladenine)",
      mod_type == "unmod" ~ "Unmodified",
      TRUE ~ mod_type
    ),
    full_label = paste0(base, " - ", mod_label),
    # Create age categories
    age_category = case_when(
      Year.Collected >= 2000 ~ "Modern (<25 years)",
      Year.Collected >= 1960 ~ "Mid-age (1960s)",
      TRUE ~ "Historic (>100 years)"
    )
  )
```

# Modifications vs. Sample Age

```{r plot-by-age, fig.height=9}
p_age <- plot_data_long %>%
  filter(mod_type != "unmod") %>%
  ggplot(aes(x = Year.Collected, y = percentage, color = base)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
  facet_wrap(~mod_label, scales = "free_y", ncol = 1) +
  scale_color_manual(
    values = c("C" = "#E41A1C", "A" = "#377EB8"),
    name = "Base"
  ) +
  labs(
    title = "DNA Modifications by Sample Age",
    subtitle = "Linear regression with 95% confidence interval",
    x = "Year Collected",
    y = "Percentage (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )

print(p_age)
```

# Modifications vs. DNA Integrity (DIN)

The DNA Integrity Number (DIN) is a measure of DNA quality from TapeStation analysis. Higher DIN values indicate better DNA integrity.

```{r plot-by-din, fig.height=9}
p_din <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage, color = age_category)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
  facet_wrap(~mod_label, scales = "free_y", ncol = 1) +
  scale_color_brewer(palette = "Set1", name = "Age Category") +
  labs(
    title = "DNA Modifications vs. DNA Integrity (DIN)",
    subtitle = "Linear regression with 95% confidence interval",
    x = "DNA Integrity Number (DIN)",
    y = "Percentage (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )

print(p_din)
```

# Modifications vs. Sequencing N50

N50 is a measure of read length quality - the length at which 50% of bases are in reads of that length or longer.

```{r plot-by-n50, fig.height=9}
p_n50 <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage, color = age_category)) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
  facet_wrap(~mod_label, scales = "free_y", ncol = 1) +
  scale_color_brewer(palette = "Set1", name = "Age Category") +
  labs(
    title = "DNA Modifications vs. Sequencing N50",
    subtitle = "Linear regression with 95% confidence interval",
    x = "N50 (bases)",
    y = "Percentage (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14)
  )

print(p_n50)
```

# Combined Comparison Plots

## 5mC across all three variables

```{r combined-5mc, fig.height=4, fig.width=12}
p_5mc_age <- plot_data_long %>%
  filter(mod_type == "5mC") %>%
  ggplot(aes(x = Year.Collected, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. Age", x = "Year Collected", y = "5mC (%)") +
  theme_bw()

p_5mc_din <- plot_data_long %>%
  filter(mod_type == "5mC", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. DIN", x = "DIN", y = "5mC (%)") +
  theme_bw()

p_5mc_n50 <- plot_data_long %>%
  filter(mod_type == "5mC", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. N50", x = "N50 (bases)", y = "5mC (%)") +
  theme_bw()

p_5mc_age + p_5mc_din + p_5mc_n50
```

## 5hmC across all three variables

```{r combined-5hmc, fig.height=4, fig.width=12}

p_5hmc_age <- plot_data_long %>%
  filter(mod_type == "5hmC") %>%
  ggplot(aes(x = Year.Collected, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5mC vs. Age", x = "Year Collected", y = "5hmC (%)") +
  theme_bw()

p_5hmc_din <- plot_data_long %>%
  filter(mod_type == "5hmC", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5hmC vs. DIN", x = "DIN", y = "5hmC (%)") +
  theme_bw()

p_5hmc_n50 <- plot_data_long %>%
  filter(mod_type == "5hmC", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C") +
  labs(title = "5hmC vs. N50", x = "N50 (bases)", y = "5hmC (%)") +
  theme_bw()

p_5hmc_age + p_5hmc_din + p_5hmc_n50
```



## 6mA across all three variables

```{r combined-6ma, fig.height=4, fig.width=12}
p_6ma_age <- plot_data_long %>%
  filter(mod_type == "6mA") %>%
  ggplot(aes(x = Year.Collected, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#377EB8") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8") +
  labs(title = "6mA vs. Age", x = "Year Collected", y = "6mA (%)") +
  theme_bw()

p_6ma_din <- plot_data_long %>%
  filter(mod_type == "6mA", !is.na(Tapestation.DIN)) %>%
  ggplot(aes(x = Tapestation.DIN, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#377EB8") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8") +
  labs(title = "6mA vs. DIN", x = "DIN", y = "6mA (%)") +
  theme_bw()

p_6ma_n50 <- plot_data_long %>%
  filter(mod_type == "6mA", !is.na(Estimated.N50.b)) %>%
  ggplot(aes(x = Estimated.N50.b, y = percentage)) +
  geom_point(size = 3, alpha = 0.7, color = "#377EB8") +
  geom_smooth(method = "lm", se = TRUE, color = "#377EB8") +
  labs(title = "6mA vs. N50", x = "N50 (bases)", y = "6mA (%)") +
  theme_bw()

p_6ma_age + p_6ma_din + p_6ma_n50
```

# Correlation Analysis

## Correlations with Year Collected

```{r cor-age}
cor_age <- plot_data_long %>%
  filter(mod_type != "unmod") %>%
  group_by(mod_label) %>%
  summarise(
    r = cor(Year.Collected, percentage, use = "complete.obs"),
    p_value = cor.test(Year.Collected, percentage)$p.value,
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.0001 ~ "****",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

cor_age %>%
  kable(digits = 4, caption = "Correlation with Year Collected",
        col.names = c("Modification", "r", "P-value", "N", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Correlations with DIN

```{r cor-din}
cor_din <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Tapestation.DIN)) %>%
  group_by(mod_label) %>%
  summarise(
    r = cor(Tapestation.DIN, percentage, use = "complete.obs"),
    p_value = cor.test(Tapestation.DIN, percentage)$p.value,
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.0001 ~ "****",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

cor_din %>%
  kable(digits = 4, caption = "Correlation with DIN",
        col.names = c("Modification", "r", "P-value", "N", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Correlations with N50

```{r cor-n50}
cor_n50 <- plot_data_long %>%
  filter(mod_type != "unmod", !is.na(Estimated.N50.b)) %>%
  group_by(mod_label) %>%
  summarise(
    r = cor(Estimated.N50.b, percentage, use = "complete.obs"),
    p_value = cor.test(Estimated.N50.b, percentage)$p.value,
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.0001 ~ "****",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

cor_n50 %>%
  kable(digits = 4, caption = "Correlation with N50",
        col.names = c("Modification", "r", "P-value", "N", "Sig.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Summary Statistics by Age Category

```{r summary-by-age-cat}
summary_by_age <- plot_data_long %>%
  filter(mod_type != "unmod") %>%
  group_by(age_category, mod_label) %>%
  summarise(
    n = n(),
    mean_pct = mean(percentage, na.rm = TRUE),
    sd_pct = sd(percentage, na.rm = TRUE),
    mean_din = mean(Tapestation.DIN, na.rm = TRUE),
    mean_n50 = mean(Estimated.N50.b, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(age_category, desc(mean_pct))

summary_by_age %>%
  kable(digits = 2, 
        caption = "Summary statistics by age category",
        col.names = c("Age Category", "Modification", "N", 
                      "Mean %", "SD %", "Mean DIN", "Mean N50")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  collapse_rows(columns = 1, valign = "top")
```

# Interpretation Notes

## Key Findings

- **5mC (methylcytosine)**: Standard DNA methylation mark in eukaryotes
- **5hmC (hydroxymethylcytosine)**: Oxidation product of 5mC, potentially unstable in old samples
- **6mA (N6-methyladenine)**: Uncommon in eukaryotes, may indicate:
  - Contamination (bacterial/fungal DNA)
  - Age-related degradation artifacts
  - Sequencing/basecalling artifacts in degraded DNA

## Relationships to consider

- **Age effects**: Are modifications changing with sample age?
- **DIN effects**: Does DNA integrity affect apparent modification levels?
- **N50 effects**: Does sequencing quality correlate with modification detection?

High 6mA levels in older, lower-quality samples warrant careful interpretation.

