---
title: "05-specimen-coverage"
author: "Kathleen Durkin"
date: "2025-12-15"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Create a Bash variables file

This allows usage of Bash variables across R Markdown chunks.

```{r save-bash-variables-to-rvars-file, engine='bash', eval=TRUE}
{
echo "#### Assign Variables ####"
echo ""
echo "# Data directories"
echo 'export nanopore_dir=/gscratch/srlab/kdurkin1/SIFP-nanopore'
echo 'export genome_dir=${nanopore_dir}/data/GCA_965233905.1_jaEunKnig1.1'
echo 'export reference=${genome_dir}/GCA_965233905.1_jaEunKnig1.1_genomic.fna'
echo 'export gff=${genome_dir}/GCA_965233905.1_jaEunKnig1.1_genomic.gff'
echo 'export output_dir_top=${nanopore_dir}/M-multi-group/output/05-specimen-coverage'
echo 'export merged_bam_dir=${nanopore_dir}/M-multi-group/output/04-specimen-pileup/merged_bams'
echo 'export coverage_dir=${output_dir_top}/coverage'
echo 'export metadata_csv=${nanopore_dir}/data/E_tourneforti_sequencing_composition.csv'
echo ""
echo "# Input BAM directories (one per sequencing run)"
echo 'export G1L4_bam_dir=${nanopore_dir}/A-Group1/output/06.01-G1-Library4-MinION-Dorado-recall-GPU'
echo 'export G2L2_bam_dir=${nanopore_dir}/B-Group2/output/04.01-G2-Library2-MinION-Dorado-recall-GPU'
echo 'export G2L3_bam_dir=${nanopore_dir}/B-Group2/output/05.01-G2-Library3-MinION-Dorado-recall-GPU'
echo 'export G4L1_bam_dir=${nanopore_dir}/D-Group4/output/03.01-G4-Library1-MinION-Dorado-recall-GPU'
echo 'export G4L2_bam_dir=${nanopore_dir}/D-Group4/output/04.01-G4-Library2-MinION-Dorado-recall-GPU'
echo ""
echo "# Programs"
echo 'export samtools=/srlab/programs/samtools-1.20/samtools'
echo 'export bedtools=/srlab/programs/bedtools'
echo ""
echo "# Reference genome size (E. knighti)"
echo 'export GENOME_SIZE=965233905'
echo ""
echo "# Set number of CPUs to use"
echo 'export threads=20'
echo ""
} > .bashvars
cat .bashvars
```

# Setup directories

```{r create-dirs, engine='bash', eval=TRUE}
source .bashvars

mkdir -p ${output_dir_top}
mkdir -p ${coverage_dir}
```

# View metadata

Load and display the specimen-barcode mapping metadata.

```{r view-metadata, engine='bash', eval=TRUE}
source .bashvars

echo "=== Metadata file contents ==="
cat ${metadata_csv}
```

# Calculate Coverage Statistics

Using the pre-merged specimen BAMs from `04-specimen-pileup/merged_bams/`.

## List available merged BAMs

```{r list-merged-bams, engine='bash', eval=TRUE}
source .bashvars

echo "=== Available merged BAM files ==="
ls -lh ${merged_bam_dir}/*.bam 2>/dev/null || echo "No BAM files found in ${merged_bam_dir}"

echo ""
echo "=== Index files ==="
ls -lh ${merged_bam_dir}/*.bai 2>/dev/null || echo "No index files found"
```

## Generate flagstat summaries for each specimen

`samtools flagstat` provides basic alignment statistics including total reads, mapped reads, etc.

```{r flagstat-all, engine='bash', eval=TRUE}
source .bashvars

echo "=== Generating flagstat summaries for all specimen BAMs ==="

# Create output file with header
flagstat_output="${coverage_dir}/specimen_flagstat_summary.tsv"
echo -e "specimen_id\ttotal_reads\tmapped_reads\tmapped_percent\tproperly_paired\tsingletons" > ${flagstat_output}

# Process each merged BAM
for bam in ${merged_bam_dir}/*.bam; do
    if [ -f "$bam" ]; then
        # Extract specimen ID from filename (assumes format: specimen_XXXXX_merged.bam or similar)
        specimen_id=$(basename "$bam" | sed 's/_merged.bam//' | sed 's/.bam//')
        
        echo "Processing: ${specimen_id}"
        
        # Run flagstat and parse output
        flagstat_result=$($samtools flagstat -@ ${threads} "$bam")
        
        # Parse key metrics
        total=$(echo "$flagstat_result" | grep "in total" | awk '{print $1}')
        mapped=$(echo "$flagstat_result" | grep "mapped (" | head -1 | awk '{print $1}')
        mapped_pct=$(echo "$flagstat_result" | grep "mapped (" | head -1 | grep -oP '\(\K[0-9.]+')
        paired=$(echo "$flagstat_result" | grep "properly paired" | awk '{print $1}')
        singletons=$(echo "$flagstat_result" | grep "singletons" | awk '{print $1}')
        
        # Handle missing values
        paired=${paired:-0}
        singletons=${singletons:-0}
        
        # Append to output
        echo -e "${specimen_id}\t${total}\t${mapped}\t${mapped_pct}\t${paired}\t${singletons}" >> ${flagstat_output}
    fi
done

echo ""
echo "=== Flagstat Summary ==="
column -t -s$'\t' ${flagstat_output}
```

Keep in mind, these BAM files have already been filtered for mapped reads

## Generate detailed stats with samtools stats

`samtools stats` provides more detailed statistics including bases, read lengths, etc.

```{r samtools-stats, engine='bash', eval=TRUE}
source .bashvars

echo "=== Generating detailed stats for all specimen BAMs ==="

# Create output file with header
stats_output="${coverage_dir}/specimen_stats_summary.tsv"
echo -e "specimen_id\ttotal_reads\ttotal_bases\tmapped_reads\tmapped_bases\tavg_read_length\tmax_read_length\tavg_quality" > ${stats_output}

# Process each merged BAM
for bam in ${merged_bam_dir}/*.bam; do
    if [ -f "$bam" ]; then
        specimen_id=$(basename "$bam" | sed 's/_merged.bam//' | sed 's/.bam//')
        
        echo "Processing: ${specimen_id}"
        
        # Run samtools stats and extract key metrics
        stats_result=$($samtools stats -@ ${threads} "$bam" | grep "^SN")
        
        # Parse metrics (field 3 contains the value)
        total_reads=$(echo "$stats_result" | grep "raw total sequences:" | awk '{print $4}')
        total_bases=$(echo "$stats_result" | grep "total length:" | awk '{print $4}')
        mapped_reads=$(echo "$stats_result" | grep "reads mapped:" | awk '{print $4}')
        mapped_bases=$(echo "$stats_result" | grep "bases mapped:" | awk '{print $4}')
        avg_length=$(echo "$stats_result" | grep "average length:" | awk '{print $4}')
        max_length=$(echo "$stats_result" | grep "maximum length:" | awk '{print $4}')
        avg_qual=$(echo "$stats_result" | grep "average quality:" | awk '{print $4}')
        
        # Append to output
        echo -e "${specimen_id}\t${total_reads}\t${total_bases}\t${mapped_reads}\t${mapped_bases}\t${avg_length}\t${max_length}\t${avg_qual}" >> ${stats_output}
    fi
done

echo ""
echo "=== Detailed Stats Summary ==="
column -t -s$'\t' ${stats_output}
```

```
=== Detailed Stats Summary ===
specimen_id   total_reads  total_bases  mapped_reads  mapped_bases  avg_read_length  max_read_length  avg_quality
USNM_14399    sequences:   253380552    980586        253380552     258              292176           27.9
USNM_1606826  sequences:   816444871    1579542       816444871     517              405812           30.2
USNM_1740336  sequences:   835964905    1856788       835964905     450              635194           29.9
USNM_1740363  sequences:   1806427349   3244592       1806427349    557              1350671          30.3
USNM_42137    sequences:   81899916     345250        81899916      237              501909           26.3
USNM_50368    sequences:   75603529     312720        75603529      242              351578           26.3
USNM_51732    sequences:   437353191    1311434       437353191     333              781018           32.0
USNM_51861    sequences:   373493752    1267539       373493752     295              604368           29.1
USNM_51892    sequences:   485476690    1687554       485476690     288              772559           29.4
```

## Calculate coverage per specimen

Calculate genome-wide coverage using the mapped bases and genome size.

```{r calculate-coverage, engine='bash', eval=TRUE}
source .bashvars

echo "=== Calculating coverage statistics ==="

# Create comprehensive coverage output
coverage_output="${coverage_dir}/specimen_coverage_summary.tsv"
echo -e "specimen_id\ttotal_reads\tmapped_reads\tmapping_rate\ttotal_bases\tmapped_bases\tmean_read_length\testimated_coverage_X" > ${coverage_output}

# Process each merged BAM
for bam in ${merged_bam_dir}/*.bam; do
    if [ -f "$bam" ]; then
        specimen_id=$(basename "$bam" | sed 's/_merged.bam//' | sed 's/.bam//')
        
        echo "Processing: ${specimen_id}"
        
        # Get stats
        stats_result=$($samtools stats -@ ${threads} "$bam" | grep "^SN")
        
        total_reads=$(echo "$stats_result" | grep "raw total sequences:" | awk '{print $4}')
        mapped_bases=$(echo "$stats_result" | grep "bases mapped (cigar):" | awk '{print $5}')
        total_bases=$(echo "$stats_result" | grep "total length:" | awk '{print $4}')
        mapped_bases=$(echo "$stats_result" | grep "bases mapped:" | awk '{print $4}')
        avg_length=$(echo "$stats_result" | grep "average length:" | awk '{print $4}')
        
        # Calculate mapping rate
        if [ "$total_reads" -gt 0 ]; then
            mapping_rate=$(echo "scale=4; $mapped_reads / $total_reads * 100" | bc)
        else
            mapping_rate=0
        fi
        
        # Calculate coverage (mapped bases / genome size)
        if [ "$GENOME_SIZE" -gt 0 ]; then
            coverage=$(echo "scale=4; $mapped_bases / $GENOME_SIZE" | bc)
        else
            coverage=0
        fi
        
        # Append to output
        echo -e "${specimen_id}\t${total_reads}\t${mapped_reads}\t${mapping_rate}\t${total_bases}\t${mapped_bases}\t${avg_length}\t${coverage}" >> ${coverage_output}
    fi
done

echo ""
echo "=== Coverage Summary (Reference genome: ${GENOME_SIZE} bp) ==="
column -t -s$'\t' ${coverage_output}
```

```
=== Coverage Summary (Reference genome: 965233905 bp) ===
specimen_id   total_reads  mapped_reads  mapping_rate  total_bases  mapped_bases  mean_read_length  estimated_coverage_X
USNM_14399    sequences:   980586        0             253380552    253380552     258               
USNM_1606826  sequences:   1579542       0             816444871    816444871     517               
USNM_1740336  sequences:   1856788       0             835964905    835964905     450               
USNM_1740363  sequences:   3244592       0             1806427349   1806427349    557
```

## Generate per-chromosome coverage (optional)

Use `samtools coverage` for per-chromosome/contig coverage statistics.

```{r samtools-coverage, engine='bash', eval=TRUE}
source .bashvars

echo "=== Generating per-chromosome coverage ==="

# Create output directory for per-specimen coverage files
mkdir -p ${coverage_dir}/per_chromosome

for bam in ${merged_bam_dir}/*.bam; do
    if [ -f "$bam" ]; then
        specimen_id=$(basename "$bam" | sed 's/_merged.bam//' | sed 's/.bam//')
        
        echo "Processing: ${specimen_id}"
        
        # Generate coverage per chromosome/contig
        $samtools coverage "$bam" > "${coverage_dir}/per_chromosome/${specimen_id}_coverage.tsv"
        
        # Show summary (top 10 contigs by coverage)
        echo "Top 10 contigs by coverage for ${specimen_id}:"
        head -1 "${coverage_dir}/per_chromosome/${specimen_id}_coverage.tsv"
        tail -n +2 "${coverage_dir}/per_chromosome/${specimen_id}_coverage.tsv" | sort -t$'\t' -k6 -rn | head -10
        echo ""
    fi
done
```

## Generate depth histogram (optional)

Calculate depth distribution across the genome.

```{r depth-histogram, engine='bash', eval=TRUE}
source .bashvars

echo "=== Generating depth histograms ==="

mkdir -p ${coverage_dir}/depth_histograms

for bam in ${merged_bam_dir}/*.bam; do
    if [ -f "$bam" ]; then
        specimen_id=$(basename "$bam" | sed 's/_merged.bam//' | sed 's/.bam//')
        
        echo "Processing: ${specimen_id}"
        
        # Generate depth histogram using samtools depth
        # This can be slow for large files, so we'll sample
        $samtools depth -a "$bam" | \
            awk '{depth[$3]++} END {for (d in depth) print d, depth[d]}' | \
            sort -n > "${coverage_dir}/depth_histograms/${specimen_id}_depth_hist.tsv"
        
        # Summary statistics
        echo "Depth distribution summary for ${specimen_id}:"
        $samtools depth "$bam" | \
            awk '{sum+=$3; if($3>0) covered++; total++} 
                 END {
                     print "Total positions:", total; 
                     print "Covered positions:", covered; 
                     print "Mean depth:", sum/total;
                     print "Breadth of coverage:", covered/total*100, "%"
                 }'
        echo ""
    fi
done
```

# Combine with Metadata

Merge coverage statistics with specimen metadata for downstream analysis.

```{r merge-with-metadata, eval=TRUE}
library(tidyverse)

# Read metadata
metadata <- read_csv("../data/E_tourneforti_sequencing_composition.csv")

# Read coverage summary (adjust path as needed)
coverage <- read_tsv("../output/05-specimen-coverage/coverage/specimen_coverage_summary.tsv")

# Extract catalog number from specimen_id (assuming format includes catalog number)
coverage <- coverage %>%
  mutate(
    # Adjust this regex based on your actual specimen_id format
    Catalog.Number = as.numeric(str_extract(specimen_id, "\\d+"))
  )

# Join with metadata
coverage_with_metadata <- coverage %>%
  left_join(
    metadata %>% 
      select(Catalog.Number, Year.Collected, Precise.Locality, Group, Library) %>%
      distinct(),
    by = "Catalog.Number"
  ) %>%
  # Add age group classification

  mutate(
    age_group = case_when(
      Year.Collected >= 2000 ~ "Modern (2000s)",
      Year.Collected >= 1950 ~ "Mid-age (1960s)",
      TRUE ~ "Old (1880s-1910s)"
    )
  )

# Display summary
coverage_with_metadata %>%
  select(specimen_id, Catalog.Number, Year.Collected, age_group, 
         mapped_reads, mapping_rate, estimated_coverage_X) %>%
  arrange(desc(estimated_coverage_X)) %>%
  knitr::kable(caption = "Coverage Summary by Specimen")
```

# Summary Visualizations

```{r coverage-plots, eval=TRUE, fig.width=10, fig.height=6}
library(ggplot2)

# Coverage by specimen
p1 <- ggplot(coverage_with_metadata, 
             aes(x = reorder(specimen_id, estimated_coverage_X), 
                 y = estimated_coverage_X,
                 fill = age_group)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Sequencing Coverage by Specimen",
    subtitle = "E. tourneforti - Mapped reads aligned to E. knighti reference",
    x = "Specimen ID",
    y = "Coverage (X)",
    fill = "Age Group"
  ) +
  scale_fill_manual(values = c(
    "Modern (2000s)" = "#27ae60",
    "Mid-age (1960s)" = "#3498db",
    "Old (1880s-1910s)" = "#e74c3c"
  )) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 9))

print(p1)

# Save plot
ggsave("../output/05-specimen-coverage/coverage_by_specimen.png", 
       p1, width = 10, height = 8, dpi = 150)
```

```{r mapping-rate-plot, eval=TRUE, fig.width=10, fig.height=6}
# Mapping rate by specimen
p2 <- ggplot(coverage_with_metadata, 
             aes(x = reorder(specimen_id, mapping_rate), 
                 y = mapping_rate,
                 fill = age_group)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
  labs(
    title = "Mapping Rate by Specimen",
    x = "Specimen ID",
    y = "Mapping Rate (%)",
    fill = "Age Group"
  ) +
  scale_fill_manual(values = c(
    "Modern (2000s)" = "#27ae60",
    "Mid-age (1960s)" = "#3498db",
    "Old (1880s-1910s)" = "#e74c3c"
  )) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 9))

print(p2)

ggsave("../output/05-specimen-coverage/mapping_rate_by_specimen.png", 
       p2, width = 10, height = 6, dpi = 150)
```

```{r summary-by-age-group, eval=TRUE}
# Summary statistics by age group
age_summary <- coverage_with_metadata %>%
  group_by(age_group) %>%
  summarise(
    n_specimens = n(),
    mean_coverage = mean(estimated_coverage_X, na.rm = TRUE),
    sd_coverage = sd(estimated_coverage_X, na.rm = TRUE),
    median_coverage = median(estimated_coverage_X, na.rm = TRUE),
    mean_mapping_rate = mean(mapping_rate, na.rm = TRUE),
    total_mapped_reads = sum(mapped_reads, na.rm = TRUE),
    .groups = "drop"
  )

age_summary %>%
  knitr::kable(caption = "Coverage Summary by Age Group", digits = 2)
```

# Save Final Outputs

```{r save-outputs, engine='bash', eval=TRUE}
source .bashvars

echo "=== Output files ==="
ls -lh ${coverage_dir}/

echo ""
echo "=== Coverage summary file ==="
cat ${coverage_dir}/specimen_coverage_summary.tsv
```

# Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
